The role of habitat quality in host metapopulations exposed to environmentally persistent pathogens. Or Identifying when environmentally persistent pathogens transform good habitat into ecological traps

### 2015-06-15 Manuscript/manuscript.tex
a few more edits to the discussion

This work demonstrates that the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.  
These impacts are driven largely by the interaction between pathogen traits (i.e. disease-induced mortality, environmental and direct transmission rates) and the colonization and extinction rates of high and low quality habitat.

The findings that low quality habitat can inhibit disease spread while high quality habitat can enhance it suggest the presence of a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
In this light, this work expands on previous studies of disease spread in a homogeneous metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).  
Shifting the patch quality distribution towards high quality habitat in our model is  conceptually similar to increasing movement in these models.
And, as in these models, the net consequences of this trade-off for the host metapopulation depend on the traits of the pathogen.
Broadly, we can divide the observed dynamics into three different scenarios: pathogen traits produce very weak pathogen spread and high quality habitat is a net benefit; pathogen traits facilitate intermediate/endemic pathogen spread and low quality habitat is a net benefit; and pathogen traits produce easy/pandemic pathogen spread and high quality habitat is again a net benefit.

@@

Thus, if high quality habitat is preferred by migrants (i.e. $\xi_{em} > 0$), it becomes an ecological trap, that is, preferentially selected habitat that reduces fitness relative to other habitats (\cite{Robertson2006}).
This fitness cost (i.e. the disease risk) becomes more severe as pathogen longevity increases, reflecting the more prominent role of environmental transmission on high quality habitat.
Moreover, the preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct transmission from infectious migrants also selecting high quality habitat.
This echoes results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

@@

Thus, for pathogens with traits that facilitate this intermediate level of spread, we see a reversal of the relative importance of high and low quality habitat for overall host metapopulation stability.
Here, rather than serving as sources that increase host population size, we see that high quality patches become sinks.  
More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  
Indeed, we see that the combination of the trap effect (i.e. host preference for high quality habitat) along with an environmentally long-lived pathogen can have severe consequences for metapopulation stability, substantially increasing the risk of host extinction (Fig~\ref{pext}).
From a management perspective, it is critical to understand this shift in the effect of habitat quality, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (\cite{Strasser2010}). 
Generally, this means that managers should focus on maintaining high quality habitat, but Strasser et al. (2010) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
Similarly, our model suggests that in some cases, low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can maintain larger total host populations than high quality distributions.

@@

However, we find that as the environmental longevity of the pathogen increases, the region in which we observe this reversal becomes smaller, that is, high quality habitat is a net benefit for a larger range of potential pathogens (Fig~\ref{highvlow}, red region grows from panels a. to c.).
Environmentally long-lived pathogens are able to spread widely for a larger range of conditions, largely overwhelming the ability of low quality habitat to provide refuges for susceptible hosts. 
These dynamics reflect the findings of \cite{Gog2002}, who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.  
Park (\cite{Park2012}) expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was again a net benefit to the host metapopulation.  
These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g. from long-lived environmental reservoirs), factors that facilitate metapopulation stability (e.g. increased movement, high quality habitat) are generally a greater benefit than factors that inhibit pathogen spread (e.g. decreased movement, low quality habitat).

>>>>

This work demonstrates that (PCC: distribution may often be confused with spatial distribution rather than quality) habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population. (PCC: deleted, lots of keywords but no information).  In many cases, high quality habitat can enhance the disease spread and impact of environmentally persistent pathogens.  In addition, the habitats with many resources may become ecological traps that attract hosts, but are a net drain on the metapopulation due to the impacts of disease in those patches. This suggests a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability (PCC: rather than saying "processes", elaborate on what specific process you mean. ).

This work expands on previous studies of disease spread in a homogeneous metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).  
Shifting the patch quality distribution towards high quality habitat in our model is conceptually similar to increasing movement in these models.
And, as in these models, the net consequences of this trade-off for the host metapopulation depend on the traits of the pathogen.
Broadly, we can divide the observed dynamics into three different scenarios: limited pathogen spread, intermediate, and high pathogen endemicity (PCC: shouldn't there be a fourth scenario of pathogen drives the host extinct?). High quality habitats are a net benefit when pathogen spread is limited or is high enough that high and low quality patches are affected by disease. There is an intermediate scenario, however, where low quality habitats are more likely to remain uninfected and as a result boost the metapopulation size while high quality patches that are heavily infected tend to reduce the metapopulation size. (PCC Note: most things you label as pathogen traits are actually emergent properties of the host-pathogen-environment...e.g. virulence)  (PCC: my text here makes some of the stuff below redundant)

@@

Thus, if high quality habitat is preferred by migrants (i.e. $\xi_{em} > 0$), it becomes an ecological trap, that is, preferentially selected habitat that reduces fitness relative to other habitats (\cite{Robertson2006}).
This fitness cost (i.e. the disease risk) becomes more severe as pathogen longevity increases, reflecting the more prominent role of environmental transmission on high quality habitat. (PCC: You're burying the lead. This stuff should go to the beginning of the discussion. )
Moreover, the preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct transmission from infectious migrants also selecting high quality habitat.
This echoes results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

@@

Thus, for pathogens with traits that facilitate this intermediate level of spread, we see a reversal of the relative importance of high and low quality habitat for overall host metapopulation stability.
Here, rather than serving as sources that increase host population size, we see that high quality patches become sinks.  
More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  
Indeed, we see that the combination of the trap effect (i.e. host preference for high quality habitat) along with an environmentally long-lived pathogen can have severe consequences for metapopulation stability, substantially increasing the risk of host extinction (Fig~\ref{pext}).
From a management perspective, it is critical to understand this shift in the effect of habitat quality, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (\cite{Strasser2010}). 
Generally, this means that managers should focus on maintaining high quality habitat, but Strasser et al. (2010) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
Similarly, our model suggests that in some cases, low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can maintain larger total host populations than high quality distributions.  (PCC: you could also relate to the early metapopulation work, which suggested how even unoccupied habitats were important to metapopulation persistence.  Here we are layering on environmental pathogens as another reason why lower quality habitat and/or unoccupied habitats are important for conservation efforts. This also relates to migration-disease work, whereby migration can help reduce disease by moving the host off of contaminated sites (although there are also other mechanisms)

@@

However, we find that as the environmental longevity of the pathogen increases, the region in which we observe this reversal becomes smaller, that is, high quality habitat is a net benefit for a larger range of potential pathogens (Fig~\ref{highvlow}, red region grows from panels a. to c.).  (PCC: would it be more accurate to describe the effect of envi. longevity as unimodal whereby at low longevities the pathogen works as though it were directly transmitted...and so there is a strong correlation between disease occupancy and host occupancy of a patch.  At intermediate longevity, low quality patches are important as they are less likely to be occupied allowing time for decontamination. At high levels of longevity, patches are not unoccupied for long enough for decontamination to occur.  If this is a fair representation, then we should avoid saying things like the 1st sentence as it implied a monotonic relationship. )
Environmentally long-lived pathogens are able to spread widely for a larger range of conditions, largely overwhelming the ability of low quality habitat to provide refuges for susceptible hosts. 
These dynamics reflect the findings of \cite{Gog2002}, who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.  
Park (\cite{Park2012}) expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was again a net benefit to the host metapopulation.  
These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g. from long-lived environmental reservoirs), factors that facilitate metapopulation stability (e.g. increased movement, high quality habitat) are generally a greater benefit than factors that inhibit pathogen spread (e.g. decreased movement, low quality habitat).
